### 2016-06-20 motivation.tex
edits to op sem

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it\footnote{For clarity, the effects of different
  transactions are shown in different colored backgrounds.} In the
first execution, transaction {\bf Wd1} reads the current balance
(\C{k}) and writes the new balance (\C{k-a1}), but before it commits
transaction {\bf Wd2} executes and commits, writing the new balance
(\C{k-a2}). RC isolation prevent {\bf Wd2} from witnessing the
uncommitted writes of transaction {\bf Wd1}.  Subsequently committing
{\bf Wd1} leads to the loss of {\bf Wd2}'s updates (the so called
\emph{lost update anomaly}~\cite{berenson}), resulting in an incorrect
balance of \C{k-a1}. The second execution describes a similar scenario
with {\bf Wd1} and {\bf Wd2} exchanging their roles.  Clearly, {\sc
  rc} is an excessively weak isolation level for this program because
it loses the updates of one transaction, resulting in the violation of
the post-condition.  A stronger isolation level that prevents lost
updates is required.  \iso{Snapshot Isolation} ({\sc
  si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} is not functionally the same as
  {\sc ser}, however; the differences in the two manifest in the \emph{absence}
  of any write-write conflicts.}   Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based
concurrency control, it is also more efficient, making it appropriate
appropriate for both {\bf Wd} transactions.

@@

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}. Rely-guarantee is a compositional
proof technique that allows us to reason about one thread at a time by
abstracting away interference induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under weak isolation, determining what
constitutes an interference and what does not is a non-trivial
problem. For example, inside a serializable transaction there is no
interference, wherase inside a transaction executing under \iso{Read
  Uncommitted}, the weakest isolation level, all interferences are
valid. Between these two extremes are various levels of isolation that
admit some interferences while prohibiting others.
% For example, \iso{Read Committed} isolation admits interference of a
% committed transaction, but not that of an uncommitted transaction.
% \iso{Snapshot Isolation} admits interference from committed
% non-conflicting transactions, and so on. 
For the rely-guarantee approach to be useful, it has to thus constrain the
interference in accordance with the chosen level of isolation. The
first key observation we make is that we can constrain interference by
constraining the definition of the \emph{happens-before} ($\hbZ$) relation
between transactions, or their respective constituents.  To do so, we
axiomatize  the $\hbZ$ relation to capture the interference characteristics of
different isolation levels as their specifications ($\psi$). For example,
a \iso{Read Committed} specification allows an operation $\eta_1$ of
$T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
operation in $T_1$ (including its commit) happens before $\eta_2$ (we
write $T_1 \hboar \eta_2$ to denote this). The {\sc rc} specification does not require
$T_1$ to execute and commit before all operations of $T_2$, thus allowing it to
interfere in $T_2$. The  \iso{Snapshot
Isolation} specification requires two transactions that write to the same
variable to be related by happens-before.  This effectively prohibits
interference due to actions in $T_1$ being interleaved in $T_2$ (or, vise versa).
Formally\footnote{Note that specifications are simplified for illustration. The
actual specifications described in (\S~\ref{sec:opsem}) are more nuanced.  }:
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall \eta_1,\eta_2,T_1,T_2.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
from {\bf Wd2} to {\bf Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
arrows are legal under $\psi_{RC}$ because in no case does an operation
$\eta_1$ from {\bf Wd1} happen-before $\eta_2$ of {\bf Wd2} without the commit
of {\bf Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ between actions in {\bf Wd2}
and only a subset of the operations in {\bf Wd1}. Ligher (blue) arrows with
hollow heads denote $\hbZ$ relationships that do not affect the value
of \C{C} in a way that causes the program to violate its
post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
relationships that lead to the violation of the post-condition. The task of
the reasoning framework is to determine if all $\hbZ$ relationships
allowed by an isolation level lead to the satisfaction of the
post-condition.

@@

Specification of an isolation level encodes its interference
characteristic as constraints over the $\hbZ$ relation. However, for this
to be useful in reasoning about programs, the rely-guarantee framework
should be able to use the specification to determine if an
interference is valid or invalid, allowing the programmer to only
focus on the former. Our second key observation is that this is
possible if the reasoning framework adequately tracks $\hbZ$ at each
program point, while preserving $\hbZ$ constraints as invariants
between program points. An interference that leads to the violation of
$\hbZ$ constraints (i.e., an invalid interference) is thus
automatically prohibited. For instance, consider the program point
after the write to \C{C} in {\bf Wd1}. The expected invariant ($\phi$) at
that program point is shown below ($\committed$ stands for
``committed''):
\begin{smathpar}
\begin{array}{l}
  \neg\committed(\C{Wd1}) \conj (\neg\committed(\C{Wd2}) \Rightarrow
  \C{C = k-a1}) 
                \\
      \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
                \Rightarrow \C{C = k-a1-a2})\}
\end{array}
\end{smathpar}
$\phi$ asserts that {\bf Wd1} is not yet committed, and that it wrote to
\C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}
depending on whether or not {\bf Wd2} is committed. If $\phi$ remains
invariant until {\bf Wd1} commits, then the post-condition (\C{C =
k-a1-a2}) can be established easily. However, an interference from
{\bf Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a1}$ to \C{C} and committing {\bf Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from {\bf Wd2} starting from a
execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds (former does not hold because at least one write
from {\bf Wd1} has happened before at least one operation of {\bf Wd2}, and
the latter does not hold because {\bf Wd1} has not yet committed, and {\bf Wd2}
has already begun). Since {\bf Wd2} also writes to \C{C}, this violates
the $\psi_{SI}$ constraint which we assume to be an invariant. A proof
that the post-condition holds now follows from the contradiction. It is
informative to note that if the invariant is $\psi_{RC}$ instead of
$\psi_{SI}$, we cannot derive a contradiction and we cannot rule out
the interference, which (rightfully) causes  the proof to fail.

We have thus far assumed a strongly consistent ({\sc SC}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behavior of {\sc SC} to totally
order all operations w.r.t the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately and perennially visible to subsequent operations. This
semantics can be built into the reasoning framework leading to a proof
system tailor-made for such stores. However, for the reasoning
framework to be truly useful, it has to be parametric over different
store consistency semantics, supporting stores that are weaker than
{\sc sc} (e.g., eventually consistent), and should be able to
reconcile conflicts between consistency and isolation
constraints.  We demonstrate how our reasoning framework
makes this possible in the following sections.

>>>>

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\caption{A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  a read committed isolation level.}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it\footnote{For clarity, the effects of different
  transactions are shown in different colored backgrounds.} In the
first execution, transaction {\bf Wd1} reads the current balance
(\C{k}) and writes the new balance (\C{k-a1}), but before it commits
transaction {\bf Wd2} executes and commits, writing the new balance
(\C{k-a2}). RC isolation prevent {\bf Wd2} from witnessing the
uncommitted writes of transaction {\bf Wd1}.  Subsequently committing
{\bf Wd1} leads to the loss of {\bf Wd2}'s updates (the so called
\emph{lost update anomaly}~\cite{berenson}), resulting in an incorrect
balance of \C{k-a1}. The second execution describes a similar scenario
with {\bf Wd1} and {\bf Wd2} exchanging their roles.  Clearly, {\sc
  rc} is an excessively weak isolation level for this program because
it loses the updates of one transaction, resulting in the violation of
the post-condition.  A stronger isolation level that prevents lost
updates is required.  \iso{Snapshot Isolation} ({\sc
  si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} is not functionally the same as
  {\sc ser}, however; the differences in the two manifest in the \emph{absence}
  of any write-write conflicts.}   Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based
concurrency control, it is also more efficient, making it appropriate
appropriate for both {\bf Wd} transactions.

@@

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about one thread at a time by abstracting away
interference induced by other threads (collectively called \emph{the
  environment}) into a \emph{rely} relation. In an ordinary concurrent
program, every environment step is a valid interference in the current
thread. However, in the presence of transactions executing under weak
isolation, determining what constitutes an interference is a
non-trivial problem. For example, inside a serializable transaction
there is no interference, whereas within a transaction executing under
\iso{Read Uncommitted}, the weakest isolation level, all interferences
are valid. Between these two extremes are various levels of isolation
that admit some interferences while prohibiting others.  For example,
\iso{Read Committed} isolation admits interference of a committed
transaction, but not that of an uncommitted transaction.
\iso{Snapshot Isolation} admits interference from committed
non-conflicting transactions, and so on.

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first key observation we make is
that we can constrain interference by constraining the definition of
the \emph{happens-before} ($\hbZ$) relation between transactions, or
their respective constituents.  To do so, we axiomatize the $\hbZ$
relation to capture the interference characteristics of different
isolation levels as their specifications ($\psi$). For example, a
\iso{Read Committed} specification allows an operation $\eta_1$ of
$T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
operation in $T_1$ (including its commit) happens before $\eta_2$ (we
write $T_1 \hboar \eta_2$ to denote this). The {\sc rc} specification
does not require $T_1$ to execute and commit before all operations of
$T_2$, thus allowing it to interfere in $T_2$. The \iso{Snapshot
  Isolation} specification requires two transactions that write to the
same variable to be related by happens-before.  This effectively
prohibits interference due to actions in $T_1$ being interleaved in
$T_2$ (or, vise versa).  Formally\footnote{Note that specifications
  are simplified for illustration. The actual specifications described
  in (\S~\ref{sec:opsem}) are more nuanced.  }:
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall \eta_1,\eta_2,T_1,T_2.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
from {\bf Wd2} to {\bf Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
arrows are legal under $\psi_{RC}$ because in no case does an operation
$\eta_1$ from {\bf Wd1} happen-before $\eta_2$ of {\bf Wd2} without the commit
of {\bf Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ between actions in {\bf Wd2}
and only a subset of the operations in {\bf Wd1}. Ligher (blue) arrows with
hollow heads denote $\hbZ$ relationships that do not affect the value
of \C{C} in a way that causes the program to violate its
post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
relationships that lead to the violation of the post-condition. The task of
the reasoning framework is to determine if all $\hbZ$ relationships
allowed by an isolation level lead to the satisfaction of the
post-condition.

@@

The specification of an isolation level encodes its interference
characteristic as constraints over the $\hbZ$ relation. However, for this
to be useful in reasoning about programs, the rely-guarantee framework
should be able to use the specification to determine if an
interference is valid or invalid, allowing the programmer to only
focus on the former. Our second key observation is that this is
possible if the reasoning framework adequately tracks $\hbZ$ at each
program point, while preserving $\hbZ$ constraints as invariants
between program points. An interference that leads to the violation of
$\hbZ$ constraints (i.e., an invalid interference) is thus
automatically prohibited. For instance, consider the program point
after the write to \C{C} in {\bf Wd1}. The expected invariant ($\phi$) at
that program point is shown below ($\committed$ stands for
``committed''):
\begin{smathpar}
\begin{array}{l}
  \neg\committed(\C{Wd1}) \conj (\neg\committed(\C{Wd2}) \Rightarrow
  \C{C = k-a1}) 
                \\
      \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
                \Rightarrow \C{C = k-a1-a2})\}
\end{array}
\end{smathpar}
$\phi$ asserts that {\bf Wd1} is not yet committed, and that it wrote to
\C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}
depending on whether or not {\bf Wd2} is committed. If $\phi$ remains
invariant until {\bf Wd1} commits, then the post-condition (\C{C =
k-a1-a2}) can be established easily. However, an interference from
{\bf Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a1}$ to \C{C} and committing {\bf Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from {\bf Wd2} starting from an
execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds (the former does not hold because at least one write
from {\bf Wd1} has happened before at least one operation of {\bf Wd2}, and
the latter does not hold because {\bf Wd1} has not yet committed, and {\bf Wd2}
has already begun). Since {\bf Wd2} also writes to \C{C}, this violates
the $\psi_{SI}$ constraint which we assume to be an invariant. A proof
that the post-condition holds now follows from the contradiction. It is
informative to note that if the invariant is $\psi_{RC}$ instead of
$\psi_{SI}$, we cannot derive a contradiction and we cannot rule out
the interference, which (rightfully) causes  the proof to fail.

We have thus far assumed a strongly consistent ({\sc SC}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behavior of {\sc SC} to totally
order all operations w.r.t the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately and perennially visible to subsequent operations. This
semantics can be built into the reasoning framework, leading to a proof
system tailor-made for such stores. However, for the reasoning
framework to be truly useful, it has to be parametric over different
store consistency semantics, supporting stores that are weaker than
{\sc sc} (e.g., eventually consistent), and should be able to
reconcile conflicts between consistency and isolation
constraints.  We demonstrate how our reasoning framework
makes this possible in the following sections.
